---
title: "Minimum Viable Bean Analysis"
author: "Julio Avelar-Barragan"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries and input files.  
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
library(Maaslin2)
library(RColorBrewer)
library(nlme)
library(Hmisc)

options(scipen=999)

setwd("/media/julio/Storage/BEANS/")

metadata = read.csv("bean_metadata.csv") %>% unite(col = "Fiber2", Fiber, Fraction, sep = "_", remove = F)

OTU_table = read.delim("Bracken_L7_table.txt", check.names = F, row.names = 1) %>%
  select(contains("num")) %>%
  rename_with(~gsub("_species.bracken_num", "", .x)) %>%
  select(metadata$SampleID) %>%
  filter(rowSums(.) != 0)

OTU_table = OTU_table[!rownames(OTU_table) == "Homo sapiens",]
  
Read_counts = read.delim("Bean_readcounts.tsv", check.names = F, header = T)
```

Read counts.  
```{r}
RC_plot_df = reshape2::melt(Read_counts)

grp.means = RC_plot_df %>% group_by(variable) %>% summarise(means = mean(value))
grp.means$sd = RC_plot_df %>% group_by(variable) %>% summarise(sd = sd(value))

ggplot(data = RC_plot_df) +
  aes(x = value, fill = variable) + 
  geom_vline(data=grp.means, aes(xintercept=means, color=variable), linetype="dashed") +
  geom_histogram(position = "identity", alpha = 0.5, color = "black") + 
  theme_bw() +
  labs(title = "Read counts/Sample", x = NULL, y = "Count", fill = "QC Step") + 
  guides(color = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Seems like a significant proportion of reads were lost from filtering out human, mouse, and rat derived reads.  

Kraken unknowns.  
```{r kraken_unknowns}
kraken_unknowns = read.csv("/media/julio/Storage/BEANS/kraken_report_merged_unknown.txt", sep="") %>% 
  right_join(metadata) %>%
  filter(!Fiber == "CTRL")

kraken_unknowns$Fiber = factor(kraken_unknowns$Fiber, levels = c("NONE", "BLACK", "PINTO"))

ggplot(data = kraken_unknowns) +
  aes(x = Fiber, y = Unknown, fill = Fraction) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = .1)) +
  #facet_grid(.~Time, scales = "free", space = "free") +
  theme_bw() +
  labs(x = NULL, title = "Reads not mapping to kraken2 DB", y = "% of unmapped reads")
  #scale_fill_manual(values = c("#BF0D3E", "steelblue")) +
  #theme(strip.background = element_rect(fill="white"))
```
The amount of unknowns is similar to the pure fiber culturing analysis.  

'Rarefaction' Curve.  
```{r rarefaction_curve}
rarecurve(t(OTU_table), sample = 100000, step = 10000, label = F, xlab = "Read Depth", col = "orange", xlim=c(0, 200000))

#Looks like a good idea to toss samples which had less than 100k mapped reads.  
OTU_clean = OTU_table[,!colSums(OTU_table) < 100000]
```

Alpha diversity.  
```{r Alpha}
alpha_df = as.data.frame(diversity(x = OTU_clean, MARGIN = 2)) %>% merge(., y = metadata, by.x = "row.names", by.y = "SampleID") %>% rename(Shannon = `diversity(x = OTU_clean, MARGIN = 2)`)

alpha_df = alpha_df %>% filter(!Fiber == "CTRL")
alpha_df$Fiber = factor(alpha_df$Fiber, levels = c("NONE", "BLACK", "PINTO"))

ggplot(data = alpha_df) +
  aes(x = Fiber, y = Shannon, fill = Fraction) +
  geom_boxplot(outlier.shape = NA) +
  #facet_grid(.~Time, scales = "free", space = "free") +
  geom_point(position = position_jitterdodge(jitter.width = .1), alpha = 0.5) +
  #scale_fill_manual(values = c("#BF0D3E", "steelblue")) +
  #geom_text(label = alpha_df$Row.names, position = position_jitterdodge(jitter.width = .1)) +
  theme_bw() +
  theme(strip.background = element_rect(fill="white")) +
  labs(x = NULL, y = "Shannon Index", title = "Taxonomic diversity")
```

No difference between bean or solubility types in shannon diversity.  

Beta diversity.  
```{r beta, include=FALSE}
T24_nmds = metaMDS(t(OTU_clean), trymax = 1024, parallel = 32, k = 2, )
T24_points = T24_nmds$points %>% 
  merge(., metadata, by.x = "row.names", by.y = "SampleID") %>% filter(!Fiber == "CTRL")
T24_points$Fiber = factor(T24_points$Fiber, levels = c("NONE", "BLACK", "PINTO"))
```

```{r}
ggplot(data = T24_points) +
  aes(x = MDS1, y = MDS2) +
  #stat_ellipse(aes(mapping = Fiber2, color = Fiber2), alpha = 0.5, lty = 2) +
  geom_point(aes(fill = Fraction, pch = Fiber), color = "black", size = 5, alpha = 0.5) +
  geom_text(aes(label = Subject), color = "black", size = 2.5, alpha = 0.7) +
  scale_shape_manual(values = c(21,23, 24)) +
  #scale_fill_manual(values = c("#BF0D3E", "steelblue")) +
  #scale_color_manual(values = c("#BF0D3E", "steelblue")) +
  guides(color = "none", fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  #theme(legend.position = "none") +
  #labs(title = "Time point 0h") +
  annotate("text", x = Inf, y = Inf, label = bquote("Stress ="~.(round(T24_nmds$stress, digits = 2))), hjust = 1, vjust = 1)
  #annotate("text", x = .5, y = .275, label = "k = 2", hjust = 1)
```

Permanova.  
```{r permanova}
OTU_clean_merged = as.data.frame(t(OTU_clean)) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(metadata) %>%
  filter(!Fiber == "CTRL" & !Fraction == "NONE") %>%
  column_to_rownames(var = "SampleID")

adonis2(formula = select(OTU_clean_merged, !Fiber2:Subject) ~ Subject + Fiber + Fraction, data = OTU_clean_merged, permutations = 999, method = "bray", parallel = 32)
```

Run maaslin2 for differential abundance.  
Only ran it for the soluble/insoluble fraction because that indicated significant differences by permanova.  

```{r eval=FALSE, include=FALSE}
maaslin_OTU = select(OTU_clean_merged, !Fiber2:Subject) %>% decostand(., method = "total", MARGIN =1) %>% t(.)

Maaslin2(input_data = as.data.frame(maaslin_OTU), input_metadata = select(OTU_clean_merged, Fiber2:Subject), output = "Fraction_maaslin", min_prevalence = .2, min_abundance = 0.001, normalization = "TSS", transform = "LOG", analysis_method = "LM", fixed_effects = "Fraction", max_significance = 0.05, plot_heatmap = T, plot_scatter = T, random_effects = "Subject", cores = 4)
```

Volcano plot of dabs features.  
```{r}
maas_res = read.delim("Fraction_maaslin/all_results.tsv", check.names = F)

maas_res = maas_res %>% mutate(color = case_when(
  qval < 0.05 & coef > 0 ~ "color1",
  qval < 0.05 & coef < 0 ~ "color2",
  qval > 0.05 ~ "color3"
))

ggplot(data = maas_res) +
  aes(x = coef, y = -log10(qval), color = color) +
  geom_hline(yintercept = -log10(0.05), lty = 2) +
  geom_vline(xintercept = 0, lty = 3) +
  geom_point(alpha = 0.5, size = 2) + 
  scale_color_manual(values = c("steelblue", "#BF0D3E", "gray"), labels = c("Increased", "Decreased", "N.S.")) +
  labs(title = "Significantly different microbes by fraction", subtitle = "T24 only", x = "MaAsLin coefficient", y = expression("-log"[10]*"(q-value)"), color = "Abundance in \nSoluble Fraction") +
  theme_bw() +
  ggrepel::geom_text_repel(aes(x = coef, y = -log10(qval), label = feature), color = "black", size = 2, max.overlaps = 8) +
  annotate("text", x = -1.5, y = 10.5, size = 2.5, hjust = 0, label = bquote("Total:"~.(sum(maas_res$qval < 0.05)))) +
  annotate("text", x = -1.5, y = 11.5, size = 2.5, hjust = 0, label = bquote("Positive:"~.(sum(maas_res$qval < 0.05 & maas_res$coef > 0)))) +
  annotate("text", x = -1.5, y = 11, size = 2.5, hjust = 0, label = bquote("Negative:"~.(sum(maas_res$qval < 0.05 & maas_res$coef < 0))))
```

Taxa bar plot at the family level.  
```{r fig.width=10}
family_OTU_table = read.delim("/media/julio/Storage/BEANS/Bracken_L5_table.txt", check.names = F, row.names = 1) %>%
  select(contains("frac")) %>%
  rename_with(~gsub("_family.bracken_frac", "", .x)) %>%
  select(alpha_df$Row.names) %>%
  filter(rowSums(.) != 0) %>%
  #select(!contains("NEG") & ) %>%
  rownames_to_column(var = "L5")

top_10 = family_OTU_table %>%
  rowwise(L5) %>% 
  summarise(mean = mean(c_across(`I_BK-F01-1`:`S_PN-F16-3`))) %>% 
  ungroup %>% 
  slice_max(mean, n = 10)

#Melt and separate taxonomy, remove whitespaces
barplot_df = family_OTU_table %>% 
  pivot_longer(!L5) %>%
  mutate(L5 = if_else(L5 %in% top_10$L5, L5, "Other")) %>%
  group_by(L5, name) %>%
  summarise(value = sum(value)) %>%
  mutate(x = "Family") %>% 
  left_join(metadata, by = join_by("name" == "SampleID"))

barplot_df = rbind(barplot_df[!(barplot_df$L5 == "Other"),],barplot_df[(barplot_df$L5 == "Other"),]) #Move other to bottom
barplot_df$L5 <- factor(barplot_df$L5, levels = unique(barplot_df$L5)) #Fix the order
barplot_df$Fiber <- factor(barplot_df$Fiber, levels = c("NONE", "BLACK", "PINTO"))
barplot_df$Fraction <- factor(barplot_df$Fraction, levels = c("NONE", "SOLUBLE", "INSOLUBLE"))

Julio_color = c("#003f5c", "#665191", "#d45087", "#ff7c43","#ffa600", "#7F0A57", "#CD9ABB", "#39A9AB", "#71CFC5", "#007947" ,"gray")

ggplot(data = barplot_df) +
  aes(x = name, weight = value, fill = L5) +
  geom_bar() +
  theme_classic() +
  facet_grid(.~Subject, scales = "free", space = "free") +
  labs(x = NULL, y = "Relative abundance", fill = "Family") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_fill_manual(values = Julio_color)
```